---
title: "Starting Steps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Starting Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brolgar)
```

When we first get a longitudinal dataset, you need to understand some of its structure. This vignette demonstrates part of the process of understanding your new longitudinal data.

# How many observations are there?

We can find the number of observations with `l_n_obs()`.

```{r show-nobs}
l_n_obs(wages, id)
```

A plot of this can help understand the distribution of observations

```{r plot-nobs}
library(ggplot2)
l_n_obs(wages, id) %>%
  ggplot(aes(x = l_n_obs)) + 
  geom_bar()
```

An equivalent table of the number of observations can be obtained usined `dplyr`'s `count()`.

```{r count-nobs}
l_n_obs(wages, id) %>%
  count(l_n_obs) %>%
  mutate(prop = n/sum(n)) %>%
  arrange(-prop)
```

# How many missings for each person


