---
title: "Starting Steps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Starting Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-set-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brolgar)
```

When we first get a longitudinal dataset, you need to understand some of its structure. This vignette demonstrates part of the process of understanding your new longitudinal data.

# Basic summaries of the data

# How many observations are there?

We can find the number of observations with `l_n_obs()`.

```{r show-nobs}
l_n_obs(wages, id)
```

A plot of this can help understand the distribution of observations

```{r plot-nobs}
library(ggplot2)
l_n_obs(wages, id) %>%
  ggplot(aes(x = l_n_obs)) + 
  geom_bar()
```

An equivalent table of the number of observations can be obtained usined `dplyr`'s `count()`.

```{r count-nobs}
library(dplyr)
l_n_obs(wages, id) %>%
  count(l_n_obs) %>%
  mutate(prop = n/sum(n)) %>%
  arrange(-prop)
```

# How many missings for each person

...

# Look at as much of the raw data as possible

Following a recommendation from Diggle et al.'s book, we want to look at as much of the raw data as possible.

We know that looking at the `lnw` and `expert` yields a plate of spaghetti:

```{r show-spag}
library(ggplot2)
ggplot(wages, 
       aes(x = lnw,
           y = exper,
           group = id)) + 
  geom_line()
```

We might instead want to filter down to those with >3 observations, and then look at a random set of individuals.

We can filter down to those with >3 observations using `filter_n_obs()`:

```{r filter-n-obs}
wages %>%
  filter_n_obs(id = id, l_n_obs > 3) %>%
  ggplot(aes(x = lnw,
             y = exper,
             group = id)) + 
  geom_line()
```

But that's still a lot of spaghetti.

We could then sample a set of say 100 observations, using `sample_n_obs()`:

```{r sample-n-obs}
wages %>%
  filter_n_obs(id = id, l_n_obs > 3) %>%
  sample_n_obs(id = id, size = 100) %>%
  ggplot(aes(x = lnw,
             y = exper,
             group = id)) + 
  geom_line()
```

That's not bad, but we can split this up into a few random groups with `add_k_groups`:

```{r show-add-k-groups}
wages %>%
  filter_n_obs(id = id, l_n_obs > 3) %>%
  sample_n_obs(id = id, size = 100) %>%
  add_k_groups(id = id, k = 10) %>%
  ggplot(aes(x = lnw,
             y = exper,
             group = id)) + 
  geom_line() + 
  facet_wrap(~.rand_id)
```

# Highlight patterns of interest

We can break the data into groups based on the slope of `exper~lnw`

```{r show-add-slope}
wages %>%
  filter_n_obs(id = id, l_n_obs > 10) %>%
  # sample_n_obs(id = id, size = 100) %>%
  add_l_slope(id = id, formula = exper~lnw) %>%
  mutate(slope_group = if_else(condition = l_slope_lnw > 0,
                               true = "positive slope",
                               false = "negative slope")) %>%
  ggplot(aes(x = lnw,
             y = exper,
             group = id)) + 
  geom_line() + 
  facet_wrap(~slope_group, nrow = 2)
  
```

