---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-set-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brolgar)
```

When we first get a longitudinal dataset, you need to understand some of its structure. This vignette demonstrates part of the process of understanding your new longitudinal data.

# Setting up your data

To use `brolgar` with your work, you should convert your longitudinal data into a time series `tsibble` using the `tsibble` package. To do so, you need to identify the unique identifying `key`, and time `index`. For example:

```{r wages-ts, eval = FALSE}
library(tsibble)
wages_ts <- as_tsibble(wages_ts,
                       key = id,
                       index = xp,
                       regular = FALSE)
```

To learn more about longitudinal data as time series, see the vignette: [Longitudinal Data Structures]()

# Basic summaries of the data

When you first get a dataset, you need to get an overall sense of what is in the data.

## How many observations are there?

The number of observations can mean different things
We can kind the number of keys using `n_keys()`:

```{r n-obs}
library(tsibble)
n_keys(wages_ts)
```

Note that this is a single number, in this case, we have `r n_keys(wages_ts)` observations.

However, we might want to know how many observations we have for each individual. If we want the number of observations in each key (id), then we can use `n_key_obs`.

```{r n-key-obs}
n_key_obs(wages_ts)
```

A plot of this can help provide better understanding of the distribution of observations

```{r plot-nobs}
library(ggplot2)
n_key_obs(wages_ts) %>%
  ggplot(aes(x = n_obs)) + 
  geom_bar()
```

# Look at as much of the raw data as possible

## Can we look at a random sample of many people?

With your longitudinal data, you want to look at as much of the raw data as possible.

We know that looking at the `ln_wages` and `xp` yields a plate of spaghetti:

```{r show-spag}
library(ggplot2)
ggplot(wages_ts, 
       aes(x = xp,
           y = ln_wages,
           group = id)) + 
  geom_line()
```

We might instead want to filter down to those with >10 observations, and then look at a random set of individuals, using the `filter_n_obs()` function.

```{r}
library(tsibble)
library(dplyr)

wages_ts %>% 
  filter_n_obs(n_obs > 10)
```


```{r filter-n-obs}

wages_ts %>%
  filter_n_obs(n_obs > 10) %>%
  ggplot(aes(x = ln_wages,
             y = xp,
             group = id)) + 
  geom_line()
```


But that's still a lot of spaghetti.

We could then sample a set of say 100 observations, using `sample_n_key()`:

```{r sample-n-obs}
wages_ts %>%
  sample_n_keys(size = 100) %>%
  filter_n_obs(n_obs > 10) %>%
  ggplot(aes(x = ln_wages,
             y = xp,
             group = id)) + 
  geom_line()
```

That's not bad, but we can split this up into a few random groups with `stratify_keys`, and then plot those groups:

```{r show-add-k-groups}
wages_ts %>%
  filter_n_obs(n_obs > 3) %>%
  sample_n_keys(size = 60) %>%
  stratify_keys(n_strata = 12) %>%
  ggplot(aes(x = ln_wages,
             y = xp,
             group = id)) + 
  geom_line() + 
  facet_wrap(~.strata)
```

# What sort of time period is between measurements?

```{r}
wages_ts %>%
  features(xp, feat_ranges)
```


# What sort of time period is between each measurement?

```{r}
wages_ts %>%
  features(xp, lag)
```


# Highlight patterns of interest

We can break the data into groups based on the slope of `ln_wages ~ xp`

```{r show-add-slope, eval = FALSE}

wages_slope <- wages_ts %>%
  filter_n_obs(n_obs > 10) %>%
  add_key_slope(ln_wages ~ xp) %>%
  mutate(slope_group = if_else(condition = .slope_xp > 0,
                               true = "positive slope",
                               false = "negative slope"))

  ggplot(wages_slope,
         aes(x = xp,
             y = ln_wages,
             group = id)) + 
  geom_line() + 
  facet_wrap(~slope_group, nrow = 2)
  
```
