---
title: "Longitudinal Data Structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{longitudinal-data-structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

There are many ways to describe longitudinal data - from panel data,
cross-sectional data, and time series. We define longitudinal data as:

> individuals repeatedly measured through time.

To explore and model longitudinal data, It is important to understand what
variables represent the individual components, and the time components, and how
these identify an individual moving through time. Identifying the individual and
time components can sometimes be a challenge, so this vignette walks through how
to do this.

The tools and workflows in `brolgar` are designed to work with a special tidy
time series data frame called a `tsibble`. We can define our longitudinal data
in terms of a time series to gain access to some really useful tools. To do so,
we need to identify three components:

1.  The **key** variable in your data is the **identifier** of your individual.
2.  The **index** variable is the **time** component of your data.
3.  The **regularity** of the time interval (index). Longitudinal data typically
    has irregular time periods between measurements, but can have regular
    measurements.

Together, time **index** and **key** uniquely identify an observation.

The term `key` is used a lot in brolgar, so it is an important idea to
internalise:

> **The key is the identifier of your individuals or series**

Why care about defining longitudinal data as a time series? Once we account for this time series
structure inherent in longitudinal data, we gain access to a suite of nice tools
that simplify and accelerate how we work with time series data. `brolgar` is
built on top of the powerful [`tsibble`](https://tsibble.tidyverts.org/) package
by [Earo Wang](https://earo.me/), if you would like to learn more, see the
[official package documentation](https://tsibble.tidyverts.org/) or read [the
paper](https://arxiv.org/abs/1901.10257).

## Converting your longitudinal data to a time series

To convert longitudinal data into a "**t**ime **s**eries tibble", a
[`tsibble`](https://tsibble.tidyverts.org/), we need to consider:

  - What identifies the **time** component of the data? This is the **index**
  - What is the **unique identifier** of an individual/series? This is the
    **key**

Together, the **index** and **key** uniquely identify an observation.

What do we mean by this? Let's look at the first section of the wages, **wages**
data analysed in Singer & Willett (2003):

``` {r slice-wages}
library(brolgar)
suppressPackageStartupMessages(library(dplyr))
slice(wages, 1:10)
```

We have the `id` column, which identifies an individual.

We also have the `xp` column, which identifies the experience an individual has.

So:

  - key: `id`
  - index: `xp`

We could create a `tsibble` of this data by using the `as_tsibble` function from
`tsibble`, also stating, `regular = FALSE`, since we have an `irregular` time
series (the measurements are not regularly measured, and have a varying distance
between each measurement):

``` {r create-tsibble, eval = FALSE}
library(tsibble)
as_tsibble(x = wages,
           key = id,
           index = xp,
           regular = FALSE)
```

``` {r print-wages-tsibble, echo = FALSE}
wages
```


Note the following information printed at the top of `wages`

    # A tsibble: 6,402 x 9 [!]
    # Key:       id [888]
    ...

This says:

  - We have `r nrow(wages)` rows,
  - with `r ncol(wages)` columns.

The `!` at the top means that there is no regular spacing between series

The "key" variable is then listed - `id`, of which there `r n_keys(wages)`.

# Using time series data in `brolgar`

This idea of longitudinal data is core to brolgar. Other vignettes in the
package will further show why the time series `tsibble` is useful.

