---
title: "Longitudinal Data Structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{longitudinal-data-structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

There are many ways to describe longitudinal data - from panel data,
cross-sectional data, and time series. We define longitudinal data as:

> Information from the same individuals, recorded at multiple points in time.

To explore and model longitudinal data, It is important to understand what
variables represent the individual components, and the time components, and how
these identify an individual moving through time. Identifying the individual and
time components can sometimes be a challenge, so this vignette walks through how
to do this.


# Defining longitudinal data as a `tsibble`

The tools and workflows in `brolgar` are designed to work with a special tidy
time series data frame called a `tsibble`. We can define our longitudinal data
in terms of a time series to gain access to some really useful tools. To do so,
we need to identify three components:

1.  The **key** variable in your data is the **identifier** of your individual.
2.  The **index** variable is the **time** component of your data.
3.  The **regularity** of the time interval (index). Longitudinal data typically
    has irregular time periods between measurements, but can have regular
    measurements.

Together, time **index** and **key** uniquely identify an observation with repeated measurements

The term `key` is used a lot in brolgar, so it is an important idea to
internalise:

> **The key is the identifier of your individuals or series**

Why care about defining longitudinal data as a time series? Once we account for this time series
structure inherent in longitudinal data, we gain access to a suite of nice tools
that simplify and accelerate how we work with time series data. 

`brolgar` is
built on top of the powerful [`tsibble`](https://tsibble.tidyverts.org/) package
by [Earo Wang](https://earo.me/), if you would like to learn more, see the
[official package documentation](https://tsibble.tidyverts.org/) or read [the
paper](https://arxiv.org/abs/1901.10257).


## Converting your longitudinal data to a time series

To convert longitudinal data into a "**t**ime **s**eries tibble", a
[`tsibble`](https://tsibble.tidyverts.org/), we need to consider which variables
identify:

1.  The individual, who would have repeated measurements. This is the **key** 
2.  The time component, this is the **index** .
3.  The **regularity** of the time interval (index). 

Together, time **index** and **key** uniquely identify an observation with repeated measurements

The vignette now walks through some examples of converting longitudinal data into a `tsibble`.

# example data: wages

Let's look at the **wages** 
data analysed in Singer & Willett (2003). This data contains measurements on
hourly wages by years in the workforce, with education and race as covariates.
The population measured was male high-school dropouts, aged between 14 and 17
years when first measured. Below is the first 10 rows of the data.
 

``` {r slice-wages}
library(brolgar)
suppressPackageStartupMessages(library(dplyr))
slice(wages, 1:10) %>% knitr::kable()
```

To create a `tsibble` of the data we ask, "which variables identify...": 

1.  The **key**, the individual, who would have repeated measurements. 
2.  The **index**, the time component.
3.  The **regularity** of the time interval (index). 


Together, time **index** and **key** uniquely identify an observation with repeated measurements

From this, we can say that:

1. The **key** is the variable `id` - the subject id, from 1-888.
2. The **index** is the variable `xp` the experience in years an individual has.
3. The data is **irregular** since the experience is a fraction of year that is not an integer.

We can use this information to create a `tsibble` of this data using `as_tsibble`

``` {r create-tsibble, eval = FALSE}
library(tsibble)
as_tsibble(x = wages,
           key = id,
           index = xp,
           regular = FALSE)
```

``` {r print-wages-tsibble, echo = FALSE}
wages
```

Note that `regular = FALSE`, since we have an `irregular` time
series 

Note the following information printed at the top of `wages`

    # A tsibble: 6,402 x 9 [!]
    # Key:       id [888]
    ...

This says:

  - We have `r nrow(wages)` rows,
  - with `r ncol(wages)` columns.

The `!` at the top means that there is no regular spacing between series

The "key" variable is then listed - `id`, of which there `r n_keys(wages)`.

# example: heights data

The heights data is a little simpler than the wages data, and contains the
average male heights in 144 countries from 1810-1989, with a smaller number of
countries from 1500-1800. 

It contains four variables:

- country
- continent
- year
- height_cm

To create a `tsibble` of the data we ask, "which variables identify...": 

1.  The **key**, the individual, who would have repeated measurements. 
2.  The **index**, the time component.
3.  The **regularity** of the time interval (index). 


In this case:

- The individual is not a person, but a country
- The time is year
- The year is not regular because there are not measurements at a fixed year point.

This data is already a `tsibble` object, we can create a `tsibble` with the following code:

```{r heights-tsibble}
as_tsibble(x = heights,
           key = country,
           index = year,
           regular = FALSE)
```


# example: gapminder

The gapminder R package contains a dataset of a subset of the gapminder study (link). This contains data on life expectancy, GDP per capita, and population by country.

```{r show-gapminder}
library(gapminder)
gapminder
```

Let's identify

1.  The **key**, the individual, who would have repeated measurements. 
2.  The **index**, the time component.
3.  The **regularity** of the time interval (index). 

This is in fact very similar to the `heights` dataset:

1. The **key** is the country
2. The **index** is the year

To identify if the year is regular, we can do a bit of data exploration using `index_summary()`

```{r gap-summarise-index}
gapminder %>% 
  group_by(country) %>% 
  index_summary(year)
```

This shows us that the year is every five - so now we know that this is a regular longitudinal dataset, and can be encoded like so:

```{r tsibble-gapminder}
as_tsibble(gapminder,
           key = country,
           index = year,
           regular = TRUE)
```

# example: PISA data

The PISA study measures school students around the world on a series of math, reading, and science scores. A subset of the data looks like so:

```{r pisa-show}
pisa
```

Let's identify

1.  The **key**, the individual, who would have repeated measurements. 
2.  The **index**, the time component.
3.  The **regularity** of the time interval (index). 

Here it looks like the key is the student_id, which is nested within school_id and country,

And the index is year, so we would write the following

```r
as_tsibble(pisa, 
           key = c(country),
           index = year)
```

We can assess the regularity of the year like so:

```{r index-check}
index_regular(pisa, year)
index_summary(pisa, year)
```

We can now convert this into a `tsibble`:

```{r pisa-as-tsibble}
pisa_ts <- as_tsibble(pisa,
           key = country,
           index = year,
           regular = TRUE)

pisa_ts
```

# Conclusion

This idea of longitudinal data is core to brolgar. Understanding what longitudinal data is, and how this can be linked to a time series representation of data helps us understand our data structure, and gives us access to more flexible tools. Other vignettes in the
package will further show why the time series `tsibble` is useful.
