---
title: "Using brolgar to understand Mixed Effects Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mixed-effects-models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brolgar)
library(lme4)
library(modelr)
library(ggplot2)

```

Just as it is important to explore your data before modelling, it is important to explore your data after you fit a model, and during the modelling process.

Let's take our wages data

```{r print-wages}
wages
```

```{r wages-ggplot}
ggplot(wages,
       aes(x = xp,
           y = ln_wages,
           group = id)) + 
  geom_line()
```
Ugh. Spaghetti plot.

Let's look at a random sample of people.

```{r wages-ggplot-sample}
ggplot(wages,
       aes(x = xp,
           y = ln_wages,
           group = id)) + 
  geom_line() +
  facet_sample()
```

So let's fit a model where we look at the impact of xp and ged, and fit an intercept for each individual.

```{r fit-int}
wages_fit_int <- lmer(ln_wages ~ xp + ged + (xp |id), data = wages)
```

We can use the tools from `modelr` to add predictions and residuals to the data

```{r wages-aug}
wages_aug <- wages %>%
  add_predictions(wages_fit_int, var = "pred_int") %>%
  add_residuals(wages_fit_int, var = "res_int")
```

Now let's look at the predictions over xp

```{r gg-wages-aug}
ggplot(wages_aug,
       aes(x = xp,
           y = pred_int,
           group = id)) + 
  geom_line(alpha = 0.4) 

```

Ugh. Straight spaghetti. Let's sample that.

```{r gg-wages-sample}
ggplot(wages_aug,
       aes(x = xp,
           y = pred_int,
           group = id)) + 
  geom_line() + 
  facet_sample()
```

Now let's add in the data to the predictions.

```{r gg-wages-predictions}
wages_aug %>%
  sample_n_keys(size = 9) %>%
  ggplot(aes(x = xp,
             y = pred_int,
             group = id,
             colour = factor(id))) + 
  geom_line() + 
  geom_point(aes(x = xp,
                 y = ln_wages,
                 colour = factor(id))) + 
  facet_wrap(~id) + 
  theme(legend.position = "none")
```

And let's add some confidence bands

```{r gg-conf-band}

```


What if we grabbed a sample of those who have the best, middle, and worst residuals?

```{r summary-residuals}
summary(wages_aug$res_int)
```

We can use `keys_near()` to help with this

```{r wages-keys-near}
wages_aug_near <- wages_aug %>%
  keys_near(key = id,
            var = res_int)

wages_aug_near
```

This shows us the id's where we the residuals match closest to the five number summary. 

We can plot this data by joining it back to the wages data with predictions, to see what the spread of predictions is like.

```{r join-wages-near}
library(dplyr)
left_join(wages_aug_near,
          wages_aug,
            by = "id") %>%
  ggplot(aes(x = xp,
             y = pred_int,
             group = id,
             colour = stat)) + 
  geom_line() + 
  geom_point(aes(y = ln_wages)) + 
  facet_wrap(~id)
```


```{r gg-stratify-residuals}
# can we stratify_along to group things by the worst fits?
wages_aug %>%
  stratify_keys(n_strata = 12, 
                along = res_int) %>%
  sample_n_keys(size = 9) %>%
  ggplot(aes(x = xp,
             y = pred_int,
             group = id,
             colour = factor(id))) + 
  geom_line() + 
  geom_point(aes(x = xp,
                 y = ln_wages,
                 colour = factor(id))) + 
  facet_wrap(~.strata) + 
  theme(legend.position = "none")

```

# Exploring changes across models

We can also explore how the data match the model when we change the model

```{r}
wages_fit_int_slope <- lmer(ln_wages ~ xp + ged + (1 + xp |id), 
                            data = wages)

wages_aug_int_slope <- wages_aug %>%
  add_predictions(wages_fit_int_slope, var = "pred_int_slope") %>%
  add_residuals(wages_fit_int_slope, var = "res_int_slope")

```

# Can we keep the actual data points with the samples with facet sample?