---
title: "Stats on Stats: Workflows with Longnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stats on Stats: Workflows with Longnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r establish-chunk-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brolgar)
```

To better understand longitudinal data we can think about some of the exploratory data analysis happening over three datasets:

1. The full data, which contain profiles of the longitudinal observations
2. The longnostics, which contain the longitudinal cognostics for each id
3. The statistics on the longnostics. These are "stats on stats" of longitudinal data.

Let's expand on this.

## The full data

We have the full set of data, which we can look at in "full spaghetti" form like so:

```{r full-spaghetti}
library(ggplot2)
ggplot(wages,
       aes(x = exper,
           y = uerate,
           group = id)) + 
  geom_line()
```

Let's consider just a small piece of this data, randomly choosing a few individuals.

```{r select-rand-id}
library(dplyr)
library(tidyr)
wages %>%
  group_by(id) %>%
  nest() %>%
  sample_n(size = 10) %>% 
  unnest()

```


```{r}

```

