---
title: "Visualisation Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{visualisation-gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brolgar)
library(ggplot2)
```

# selected sample of individuals

Sample n random individuals to explore
(Note: Possibly not representative)

For example, we can sample 20 random individuals, and then plot them.

```{r selected-sample}

wages %>%
  sample_n_obs(id = id,
               size = 20)

wages %>%
  sample_n_obs(id = id,
               size = 20) %>%
  ggplot(aes(x = exper,
             y = lnw,
             group = id)) + 
  geom_line()


```

## Filter only those with certain number of observations

We can combine this with `sample_n_obs` to filter those with at least 5 observations using `filter_n_obs`. This provides us with 20 people with at least 5 observations.

```{r filter-sample}
wages %>%
  filter_n_obs(id = id,
               l_n_obs >= 5) %>%
  sample_n_obs(id = id,
               size = 20) %>%
  ggplot(aes(x = exper,
             y = lnw,
             group = id)) + 
  geom_line()
```

# Sampled facets
Assign individuals to a random group 1...K.
Facet by K

About 20 lines per plot is a good number, but we might want to split this into about 6 different groups, so we can look at many at once. So now we can look at those with at least 5 observations, then sample 120 individuals, and break them into 6 groups with `add_k_groups`. 

```{r filter-sample-group}

wages %>%
  filter_n_obs(id = id,
               l_n_obs >= 5) %>%
  sample_n_obs(id = id,
               size = 120) %>%
  add_k_groups(id = id,
               k = 6) %>%
  ggplot(aes(x = exper,
             y = lnw,
             group = id)) + 
  geom_line() + 
  facet_wrap(~.rand_id)

```

This allows for you to look at a larger set of the data.

However, it does not point you to those individuals who are "interesting", in the sense of those being outliers, or representative of the middle of the group.

# Highlight individuals (gghighlight)

We can find those individuals who have a negative slope using `add_l_slope`.
This adds columns `l_intercept` and `l_slope_exper`. Using the `gghighlight` library, we can identify those with an overall negative slope.

```{r wages-slope}

wages_slope <- wages %>%
  add_l_slope(id = id,
              formula = lnw ~ exper)

library(gghighlight)

gg_wages_slope <- ggplot(wages_slope,
       aes(x = exper,
           y = lnw,
           group = id)) + 
  geom_line() 

gg_wages_slope + 
  gghighlight(l_slope_exper < 0)

gg_wages_slope + 
  gghighlight(l_slope_exper < -0.5)

```

# Combining highlighting with animation